---
title: 'researcher-level metrics blog post'
author: "Heather Piwowar and Jason Priem"
output:
  html_document:
    fig_caption: yes
    force_captions: yes
    highlight: pygments
    keep_md: yes
    number_sections: no
    pandoc_args:
    - +RTS
    - -K10000000
    - -RTS
  md_document:
    variant: markdown_github
  pdf_document:
    fig_caption: yes
    keep_md: yes
    keep_tex: yes
    latex_engine: xelatex
---

*This paper is still in progress. Feel free to submit a pull request with updates and changes.*


```{r setup, echo=FALSE}
library(knitr)
opts_knit$set(progress=TRUE, tidy=TRUE, echo=TRUE)
```

```{r knitcitations, echo=FALSE}
  # devtools::install_github("cboettig/knitcitations@v1")
  library(knitcitations); cleanbib()
  cite_options(citation_format = "pandoc", check.entries=FALSE)
  library(bibtex)
```

```{r load_libraries, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(magrittr)
library(ggplot2)
library(dplyr)
library(broom)
library(RJDBC)
library(reshape)
library(plyr); library(dplyr)
library(scales)
library(corrplot)
```

```{r get_data, echo=FALSE, results='hide', cache=TRUE}

profiles <- read.csv("profiles.csv")
```

```{r echo=TRUE}
# subset by >25 events ad condition the columns
profiles_bool <- profiles %>% select(-id) 
# profiles_bool <- profiles %>% select(-id) %>% select(-posts) %>% select(-accounts) 
#profiles_bool[profiles_bool>0] <- 1
#profiles_bool[1,] <- 0  # hack to make corr not explode

#corrplot(cor(profiles_bool), type='lower', order='FPC', method='color')
```


how prevelant are different kinds of success for each person (by metric)
```{r}
# by metric
number_of_people = nrow(profiles)
channels = profiles %>% select(blogs:wikipedia) %>% melt
nonzero_channels =  channels %>% filter(value>1)

nonzero_channels %>% ggplot( aes(reorder(variable, variable, length)) ) +
  geom_bar(aes(y=..count../6583)) +
  coord_flip() +
  scale_y_continuous(labels = percent) +
  ggtitle("Percentage people with research mentioned on a given channel \n(people with at least 20 mentions)") +
  ylab("Percentage of people") +
  xlab("Channel")
```

how prevelant are different kinds of success for each person (overall)
```{r}
# number of sources
profiles %>% ggplot( aes(num_sources) ) + geom_histogram(binwidth=1)
profiles %>% ggplot( aes(num_sources) ) + geom_histogram(binwidth=1, aes(y=cumsum(..count..)))

# number of posts
profiles %>% ggplot( aes(posts) ) + geom_histogram()
profiles %>% ggplot( aes(posts) ) + geom_histogram() + scale_x_log10() 
profiles$posts %>% median

# altmetric scores
profiles %>% ggplot( aes(altmetric_score) ) + geom_histogram()
profiles %>% ggplot( aes(altmetric_score) ) + geom_histogram() + scale_x_log10() 
profiles$altmetric_score %>% median

# altmetric scores, zoom in on the meat
profiles %>% filter(altmetric_score < 250) %>% ggplot( aes(altmetric_score) ) + geom_histogram()

# activity, as defined by having stuff in each of the last three months
profiles %>% filter(monthly_event_count < 30) %>% ggplot( aes(monthly_event_count) ) + geom_histogram(binwidth=1)
profiles %>% filter(monthly_event_count < 30) %>% ggplot( aes(monthly_event_count) ) + geom_histogram(binwidth=1, aes(y=cumsum(..count..)))

# how many have monthly events at all
profiles %>% filter(monthly_event_count > 0) %>% nrow
profiles %>% filter(monthly_event_count == 0) %>% nrow

```





lets see how counts various metrics relate to one another

First, the number of sources vs the number of posts
```{r}
head(profiles)

# sources vs posts
profiles %>%  ggplot(aes(num_sources, posts))  + geom_jitter(alpha=0.5) 
cor(profiles$num_sources, profiles$posts)
cor(profiles$num_sources, log10(profiles$posts))
```
* moderate correlation goes up noticebly when you log-transform posts


Let's look at the relationship between the altmetric score vs posts/tweets
```{r}

# altmetric score vs posts
profiles %>%  ggplot(aes(posts, altmetric_score))  + geom_point(alpha=.2) + scale_x_log10() + scale_y_log10()
cor(log10(profiles$altmetric_score), log10(profiles$posts), use='complete.obs')
cor(profiles$altmetric_score, profiles$posts, use='complete.obs')  # makes no difference


# altmetric score vs twitter
profiles %>%  ggplot(aes(twitter, altmetric_score))  + geom_point(alpha=0.2) + scale_x_log10() + scale_y_log10()
cor(log10(profiles$altmetric_score), log10(profiles$twitter))

```

* correlations really strong, only a bit less strong when you use tweets instead of posts.


Lets take a look at the t-index
```{r}

# altmetric score vs t-index
profiles %>% ggplot( aes(t_index) ) + geom_histogram(binwidth=1, aes(y=cumsum(..count..)))
profiles$t_index %>% median


profiles %>%  ggplot(aes(t_index, altmetric_score))  + geom_jitter(alpha=0.2) + scale_x_log10() + scale_y_log10() 

cor(profiles$altmetric_score, profiles$t_index)
cor(log10(profiles$altmetric_score), log10(profiles$t_index))

```

* everything above 800 on the graph is the top 5%, which is convenient for estimating. so t-index=12 is the top 5 percentile.
* the median of 4 suggests that we could maybe do a badge for everyone who's got a 5 or better.
* surprisingly good correlation between t-index and both altmetric score and posts count, even after log-transform





